Policy optimization for dynamic spatiotemporal systems \hspace{1em} \insertframenumber

### 2014-07-31 pot/LaTeX/jsm2014.tex

\title[Spatiotemporal Modeling of Extreme Events] % (optional, use only with long paper titles)
{
  Spatiotemporal Modeling of Extreme Events
}
\author[B. Reich and S. Morris]{Brian Reich and Samuel Morris}
\institute[NCSU]{North Carolina State University}
\date{}

@@

\begin{frame}{Motivation}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Average behavior is important to understand, but it does not paint the whole picture.
    \begin{itemize}
      \item e.g. When constructing river levees, engineers need to be able to estimate a 100-year or 1000-year flood levels.
    \end{itemize}
    \item In geostatistical analysis, kriging uses spatial correlation to help inform prediction at unknown locations.
    \item Want to explore ways to incorporate this spatial correlation when estimating the tails of the distribution.
  \end{itemize}
\end{frame}

\begin{frame}{Introduction to extremes}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Max-stable processes (Cooley et al., 2012):
    \begin{itemize}
      \item Consider a spatial process $x_t(\bs)$, $t = 1, \ldots, T$.
      \item Let $M_T(\bs) = \left\{ \bigvee_{t=1}^T x_t(\bs_1), \ldots, \bigvee_{t=1}^T x_t(\bs_n) \right\}$
      \item If there exists normalizing sequences $a_T(\bs)$ and $b_T(\bs)$ such
      that for all sites, $\bs_i, i = 1, \ldots, d$,
      \begin{align*}
        a_T^{-1}(\bs) \left\{ M_T(\bs) - b_T(\bs) \right\} \converged Y(\bs)
      \end{align*}
      which has a non-degenerate distribution, then $Y(\bs)$ is a max-stable process.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{General approaches}
  \begin{itemize}
    \item Two standard approaches:
    \begin{enumerate}[1.] \setlength{\itemsep}{0.5em}
      \item Block maxima:
      \begin{itemize}
        \item Uses yearly maxima
        \item Discards many observations
      \end{itemize}
      \item Peaks-over-threshold
      \begin{itemize}
        \item Incorporates more data than block maxima
        \item Autocorrelation may be an issue between observations (e.g. flood levels don't dissipate overnight)
      \end{itemize}
    \end{enumerate}
  \end{itemize}
\end{frame}

\begin{frame}{Univariate distribution functions}
  \begin{itemize}
    \item Generalized extreme value distribution (GEV):
    \begin{align*}
      \Pr(Y_j < y) = G_j(y) = \exp \left\{ -\left[ \left(1 + \xi_j \frac{y - \mu_j}{\sigma_j}\right)_+^{-1/\xi_j} \right] \right\}
    \end{align*}
    \item Generalized Pareto distribution (GPD):
    \begin{align*}
      \Pr(Y_j > y | Y_j > T) = F_j(y) = \left( 1 + \xi_j \frac{y - T}{\sigma_j} \right)_+^{-1/\xi_j}
    \end{align*}
  \end{itemize}
\end{frame}

\begin{frame}{Multivariate representations}
  \begin{itemize}
    \item Multivariate distributions:
    \begin{itemize}
      \item Assume common standardized max-stable marginal, like unit-Fr\'{e}chet
      \begin{align*}
        \Pr(Z < z) = exp(-z^{-1})
      \end{align*}
      \item The multivariate representation for the GEV is
      \begin{align*}
        \Pr(\bZ \le \bz)  &= G^*(\bz) = \exp(-V(\bz))\\
                V(\bs)    &= d \int_{\Delta_d} \bigvee_{i = 1}^d \frac{w_i}{z_i} H(\ddd w)
      \end{align*}
      where
      \begin{itemize}
        \item $\Delta_d = \{ \bw \in \calR^d_+ \mid w_1 + \cdots + w_d = 1\}$
        \item $H$ is a probability measure on $\Delta_d$
        \item $\int_{\Delta_d}w_i H(\ddd w) = 1 / d$ for $i = 1, \ldots, d$.
      \end{itemize}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Although $V(\bs)$ can be flexible, the number of parameters is unwieldy.
    \begin{itemize}
      \item For an asymmetric logistic dependence, we have $2^d - d - 1$ free parameters
    \end{itemize}
    \item New model builds on existing geostatistical methods.
  \end{itemize}
\end{frame}

\begin{frame}{Skew-normal distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Assume observed data $Y_t(\bs)$ come from a skew-normal (Zhang and El-Shaarawi, 2012)
    \begin{align*}
      Y_t(\bs) = X_t(\bs)\beta + \alpha z_t + v_t(\bs)
    \end{align*}
    where
    \begin{itemize} \setlength{\itemsep}{0.25em}
      \item $\alpha \in \calR$ controls the skew
      \item $z_t \iid N_{(0, \infty)}(0, \sigma^2_t)$ is a random effect
      \item $v_t(\bs) \sim \text{MVN}(0, \Sigma)$ where $\Sigma_t$ is a spatial covariance matrix with variance $\sigma^2_t$
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Skew-normal distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item By conditioning on $\alpha$ and $z_t$
    \begin{align*}
      Y_t(\bs) \mid \alpha, z_t \sim \text{MVN}(\mu_t(\bs), \Sigma_t)
    \end{align*}
    where $\mu_t(\bs) = X_t(\bs) \beta + \alpha z_t$
    \item When $\sigma_t^2 \iid \text{IG}(a, b)$, then this becomes a multivariate $t$-distribution
    after marginalizing over the $\sigma_t^2$ terms.
    \item Can use standard geostatistical methods to fit this model.
    \item Predictions can be made through kriging.
  \end{itemize}
\end{frame}

\begin{frame}{Random daily partition}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Consider a set of daily knots $\{w_{t1}, \ldots, w_{tK}\}$ that define a daily partition
    $\{P_{t1}, \ldots, P_{tK}\}$ such that
    \begin{align*}
      P_{tk} = \{s : k = \argmin_\ell|| \bs - w_{t\ell}|| \}
    \end{align*}
    \item For $\bs \in P_{tk}$, the model becomes
    \begin{itemize}
      \item $\mu_t(\bs) = X_t(\bs) \beta + \alpha z_{tk}$
      \item $\sigma^2_t = \sigma^2_{tk}$
    \end{itemize}
    \item After marginalizing over the $\sigma_{tk}^2$ terms, $Y_t(\bs)$ is a multivariate $t$-distribution
    within the partition.
  \end{itemize}
\end{frame}

\begin{frame}{Thresholding data}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Tails of the distribution should speak for themselves.
    \item We threshold the observed data at a suitably high threshold $T$.
    \item Observations below the threshold are imputed during MCMC.
  \end{itemize}
\end{frame}

\begin{frame}{MCMC details}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Three main steps:
    \begin{enumerate}[1.]
      \item Impute missing observations and data below $T$
      \item Update parameters with random walk Metropolis Hastings or Gibbs sampling
      \item Make spatial predictions
    \end{enumerate}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Model comparisons}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item 9 different analysis methods
    \begin{enumerate}[1.]
      \item Gaussian
      \item $t$, $K=1$, $T=0$
      \item $t$, $K=5$, $T=0$
      \item $t$, $K=1$, $T=0.9$
      \item $t$, $K=5$, $T=0.9$
      \item skew-$t$, $K=1$, $T=0$
      \item skew-$t$, $K=1$, $T=0.9$
      \item skew-$t$, $K=5$, $T=0$
      \item skew-$t$, $K=5$, $T=0.9$
    \end{enumerate}
    \item Compare quantile scores using 5-fold cross validation
  \end{itemize}
\end{frame}

\begin{frame}{References}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Cooley, D., Cisewski, J., Erhardt, R. J., Mannshardt, E., Omolo, B. O. and Sun, Y. (2012) A survey of spatial extremes: Measuring spatial dependence and modeling spatial effects. {\it REVSTAT}, 10, 135--165.
    \item Zhang, H. and El-Shaarawi, A. (2010) On spatial skew-Gaussian processes and applications. {\it Environmetrics}, 33--47.
  \end{itemize}
\end{frame}

>>>>

\title[Spatiotemporal Modeling of Extreme Events] % (optional, use only with long paper titles)
{
  Spatiotemporal Modeling of Extreme Events
}
\author[S. Morris and B. Reich]{Samuel Morris and Brian Reich}
\institute[NCSU]{North Carolina State University}
\date{}

@@

\begin{frame}{Motivation}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Average behavior is important to understand, but it does not paint the whole picture.
    \begin{itemize}
      \item e.g. When constructing river levees, engineers need to be able to estimate a 100-year or 1000-year flood levels.
    \end{itemize}
    \item In geostatistical analysis, kriging uses spatial correlation to help inform prediction at unknown locations.
    \item Want to explore computationally easy methods that are available in higher dimensions
  \end{itemize}
\end{frame}

% \begin{frame}{Introduction to extremes}
%   \begin{itemize} \setlength{\itemsep}{0.5em}
%     \item Max-stable processes (Cooley et al., 2012):
%     \begin{itemize}
%       \item Consider a spatial process $x_t(\bs)$, $t = 1, \ldots, T$.
%       \item Let $M_T(\bs) = \left\{ \bigvee_{t=1}^T x_t(\bs_1), \ldots, \bigvee_{t=1}^T x_t(\bs_n) \right\}$
%       \item If there exists normalizing sequences $a_T(\bs)$ and $b_T(\bs)$ such
%       that for all sites, $\bs_i, i = 1, \ldots, d$,
%       \begin{align*}
%         a_T^{-1}(\bs) \left\{ M_T(\bs) - b_T(\bs) \right\} \converged Y(\bs)
%       \end{align*}
%       which has a non-degenerate distribution, then $Y(\bs)$ is a max-stable process.
%     \end{itemize}
%   \end{itemize}
% \end{frame}

\begin{frame}{Standard non-spatial analysis}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Block maxima:
    \begin{itemize}
      \item Uses yearly maxima
      \item Discards many observations
      \item Models are fit using the generalized extreme value distribution
    \end{itemize}
    \item Generalized extreme value distribution (GEV):
    \begin{align*}
      \Pr(Y_j < y) = G_j(y) = \exp \left\{ -\left[ \left(1 + \xi_j \frac{y - \mu_j}{\sigma_j}\right)_+^{-1/\xi_j} \right] \right\}
    \end{align*}
  \end{itemize}
\end{frame}

\begin{frame}{Standard non-spatial analysis}
  \begin{itemize}
    \item Peaks-over-threshold:
    \begin{itemize}
      \item Incorporates more data than block maxima
      \item Select a threshold, $T$, and fit data above the threshold using the generalized Pareto distribution
      \item Autocorrelation may be an issue between observations (e.g. flood levels don't dissipate overnight)
    \end{itemize}
    \item Generalized Pareto distribution (GPD):
    \begin{align*}
      \Pr(Y_j > y | Y_j > T) = F_j(y) = \left( 1 + \xi_j \frac{y - T}{\sigma_j} \right)_+^{-1/\xi_j}
    \end{align*}
  \end{itemize}
\end{frame}

% \begin{frame}{Multivariate representations}
%   \begin{itemize}
%     \item Multivariate distributions:
%     \begin{itemize}
%       \item Assume common standardized max-stable marginal, like unit-Fr\'{e}chet
%       \begin{align*}
%         \Pr(Z < z) = exp(-z^{-1})
%       \end{align*}
%       \item The multivariate representation for the GEV is
%       \begin{align*}
%         \Pr(\bZ \le \bz)  &= G^*(\bz) = \exp(-V(\bz))\\
%                 V(\bs)    &= d \int_{\Delta_d} \bigvee_{i = 1}^d \frac{w_i}{z_i} H(\ddd w)
%       \end{align*}
%       where
%       \begin{itemize}
%         \item $\Delta_d = \{ \bw \in \calR^d_+ \mid w_1 + \cdots + w_d = 1\}$
%         \item $H$ is a probability measure on $\Delta_d$
%         \item $\int_{\Delta_d}w_i H(\ddd w) = 1 / d$ for $i = 1, \ldots, d$.
%       \end{itemize}
%     \end{itemize}
%   \end{itemize}
% \end{frame}

\begin{frame}{Multivariate analysis}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Multivariate max-stable and GPD models have nice features, but they are
    \begin{itemize}
      \item computationally hard to work with
      \item joint distribution only available in low dimension
    \end{itemize}
    \item Pairwise likelihood approach (Huser and Davison, 2014)
  \end{itemize}
\end{frame}

\begin{frame}{Model objectives}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Our objective is to build a model that
    \begin{itemize}
      \item has a flexible tail
      \item has asymptotic spatial dependence
      \item computation on the order of Gaussian models for large space-time datasets
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Thresholding data}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item We threshold the observed data at a high threshold $T$.
    \item Thresholded data:
    \begin{align*}
      Y_t^*(\bs) = \left\{ \begin{array}{ll}
          Y_t(\bs) \quad & Y_t(\bs) > T\\
          T \quad & Y_t(\bs) \le T
      \end{array}\right.
    \end{align*}
    \item Allows tails of the distribution to speak for themselves.
  \end{itemize}
\end{frame}

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Assume observed data $Y_t(\bs)$ come from a skew-$t$ (Zhang and El-Shaarawi, 2012)
    \begin{align*}
      Y_t(\bs) = X_t(\bs)\beta + \alpha z_t + v_t(\bs)
    \end{align*}
    where
    \begin{itemize} \setlength{\itemsep}{0.25em}
      \item $\alpha \in \calR$ controls the skew
      \item $z_t \iid N_{(0, \infty)}(0, \sigma^2_t)$ is a random effect
      \item $v_t(\bs)$ is a Gaussian process with variance $\sigma^2_t$ and \Matern correlation
      \item $\sigma_t^2 \iid \text{IG}(a, b)$
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Skew-normal distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Conditioned on $z_t$ and $\sigma^2_t$, $Y_t(\bs)$ is Gaussian
    \begin{itemize}
      \item After marginalizing over $\sigma^2_t$, this becomes a multivariate $t$-distribution with $2a$ degrees of freedom.
    \end{itemize}
    \item Can use standard geostatistical methods to fit this model.
    \item Predictions can be made through kriging.
    \item Marginalizing over $z_t$ and $\sigma^2_t$ (via MCMC),
    \begin{align*}
      Y_t(\bs) \sim \text{skew-t}(\mu, \alpha, \text{df}=2a)
    \end{align*}
  \end{itemize}
\end{frame}

\begin{frame}{Long-range dependence}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item The $\chi$ coefficient is a measure of extremal spatial correlation
    \begin{align*}
      \chi(\bh) = \Pr(Y_t(\bs) > c \mid Y_t(\bs + \bh) > c)
    \end{align*}
    \item This value shows asymptotic dependence that does not approach 0 as $\bh \rightarrow \infty$.
    \item Deal with this through a daily random partition.
  \end{itemize}
\end{frame}

\begin{frame}{Random daily partition}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Daily random partition allows $z_t$ and $\sigma^2_t$ to vary by site.
    \begin{align*}
      Y_t(\bs) = X_t(\bs) \beta + \alpha z_t(\bs) + v_t(\bs)
    \end{align*}
    \item Consider a set of daily knots $\{w_{t1}, \ldots, w_{tK}\}$ that define a daily partition
    $\{P_{t1}, \ldots, P_{tK}\}$ such that
    \begin{align*}
      P_{tk} = \{s : k = \argmin_\ell|| \bs - w_{t\ell}|| \}
    \end{align*}
    \item For $\bs \in P_{tk}$
    \begin{align*}
      z_t(\bs) &= z_{tk}\\
      \sigma^2_t(\bs) &= \sigma^2_{tk}
    \end{align*}
    \item Within each partition $Y_t(\bs)$ has the same distribution as before.
  \end{itemize}
\end{frame}

\begin{frame}{MCMC details}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Three main steps:
    \begin{enumerate}[1.]
      \item Impute missing observations and data below $T$
      \item Update parameters with random walk Metropolis Hastings or Gibbs sampling
      \item Make spatial predictions
    \end{enumerate}
    \item Priors are selected to be conjugate when possible.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Model comparisons}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item 9 different analysis methods incorporating
    \begin{itemize}
      \item Gaussian vs $t$
      \item 1 partition vs 5 partitions
      \item No thresholding vs thresholded
      \begin{itemize}
        \item Thresholded data at 0.90 sample quantile
      \end{itemize}
    \end{itemize}
    \item All methods use a \Matern or exponential covariance ($\nu = 0.5$)
    \item Compare quantile and Brier scores using 5-fold cross validation
    \item Mean function modeled using a first-order spatial trend
  \end{itemize}
\end{frame}

\begin{frame}{Cross-validation results}
  \begin{table}[htbp]
    \small
    \centering
    \begin{tabular}{lrrrrrr}
           & \multicolumn{6}{c}{Quantile}\\
           \cline{2-7}
    & 0.900 & 0.950 & 0.980 & 0.990 & 0.995 & 0.999\\
  \hline
Gaussian & 16.38 & 15.76 & 15.02 & 14.52 & 14.08 & 13.22\\
$t$-1 ($T=0$) & 16.15 & 15.51 & 14.62 & 14.00 & 13.43 & 12.32\\
$t$-5 ($T=0$) & 13.61 & 12.66 & 11.61 & 10.96 & 10.40 & 9.34\\
$t$-1 ($T=0.9$) & 5.52  & 3.58  & 2.28  & 1.77  & 1.47  & 1.10\\
$t$-5 ($T=0.9$) & 5.98  & 4.27  & 2.99  & 2.41  & 2.03  & 1.49\\
skew $t$-1 ($T=0$) & 9.24  & 7.27  & 5.26  & 4.13  & 3.27  & 1.96\\
skew $t$-1 ($T=0.9$) & {\bf 4.91}  & {\bf 3.16} & {\bf 1.94}  & {\bf 1.45}  & {\bf 1.16}  & {\bf 0.82}\\
skew $t$-5 ($T=0$) & 15.81 & 14.46 & 12.74 & 11.57 & 10.57 & 8.60\\
skew $t$-5 ($T=0.9$) & & & & & &\\
\hline
    \end{tabular}
  \end{table}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Quantile scores -- Still waiting on last set
  \end{itemize}
\end{frame}

\begin{frame}{Simulation study settings}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Data generated from 6 different settings.
    \begin{enumerate}[1.]
      \item Gaussian
      \item $t$-1 with 4 degrees of freedom
      \item $t$-5 with 4 degrees of freedom
      \item skew $t$-1 with 4 degrees of freedom ($\alpha = 3$)
      \item skew $t$-5 with 4 degrees of freedom ($\alpha = 3$)
      \item Half-Gaussian, Half $t$-1 (spatial range = 0.4)
    \end{enumerate}
    \item Spatial settings
    \begin{itemize}
      \item $\bs \in [0, 1] \times [0, 1]$
      \item Exponential covariance with range: 0.1
      \item Ratio of spatial to nugget error: 0.9
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Simulation study methods}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item 5 different analysis methods
    \begin{enumerate}[1.]
      \item Gaussian
      \item Skew $t$-1 ($T=0$)
      \item Skew $t$-1 ($T=0.9$)
      \item Skew $t$-5 ($T=0$)
      \item Skew $t$-5 ($T=0.9$)
    \end{enumerate}
    \item All methods use a \Matern covariance structure except for method 5 which uses an exponential covariance ($\nu = 0.5$)
  \end{itemize}
\end{frame}

\begin{frame}{Simulation results}
  Still rerunning a few of the MCMC runs because quantile scores for 4. and 5. are
  unusually high for method 5
\end{frame}

\begin{frame}{Future work}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Comparison with extreme value analysis methods
    \item Including time in the model
    \begin{itemize}
      \item AR(1): $Y_t(\bs) = X_t(\bs) \beta + \phi Y_{t-1}(\bs) + \alpha z_t(\bs) + v_t(\bs)$
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Questions}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Any questions?
    \item Thank you for your attention.
  \end{itemize}
\end{frame}

\begin{frame}{References}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Cooley, D., Cisewski, J., Erhardt, R. J., Mannshardt, E., Omolo, B. O. and Sun, Y. (2012) A survey of spatial extremes: Measuring spatial dependence and modeling spatial effects. {\it REVSTAT}, {\bf 10}, 135--165.
    \item Huser, R. and Davison, A. C. (2014) Space-time modelling of extreme events. {\it Journal of the Royal Statistical Society: Series B (Statistical Methodology)}, {\bf 76}, 439--461
    \item Zhang, H. and El-Shaarawi, A. (2010) On spatial skew-Gaussian processes and applications. {\it Environmetrics}, {\bf 21} 33--47.
  \end{itemize}
\end{frame}
